---
title: "Covid-19 Report"
author: "Sergey Ostrovsky"
date: "5/31/2021"
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
    fig_crop: no
    highlight: haddock
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
---
# Introduction

The data Covid-19 contain region, data, number of cases and deaths, and population. My primary goal is to analyze the relationship between covid-19 cases and deaths and the percentage of cases and deaths based on population. 

# Importing Data
##### First, I will import the libraries to use for the report.
```{r}
library(tidyverse)
library(lubridate)
options(warn=-1)
```

##### Now I can load Covid-19 Data from https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/ link.

```{r}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c('time_series_covid19_confirmed_US.csv', 'time_series_covid19_confirmed_global.csv', 
                'time_series_covid19_deaths_US.csv', 'time_series_covid19_deaths_global.csv')

urls <- str_c(url_in, file_names)

US_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
US_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
```
# Tidying and Transforming Data
##### First I will transfer data rows into columns for global_cases and global_deaths tables

```{r}
global_cases <- global_cases %>%
    pivot_longer(cols = -c('Province/State',
                           `Country/Region`, Lat, Long),
                 names_to = 'date',
                 values_to = 'cases') %>%
    select(-c(Lat,Long))

global_deaths <- global_deaths %>%
    pivot_longer(cols = -c('Province/State',
                           `Country/Region`, Lat, Long),
                 names_to = 'date',
                 values_to = 'deaths') %>%
    select(-c(Lat,Long))
```

##### Next I will join global_cases an globas_deaths

```{r}
global <- global_cases %>%
    full_join(global_deaths) %>%
    rename(Country_Region = 'Country/Region',
           Province_State = 'Province/State') %>%
    mutate(date=mdy(date))

summary(global)
```

##### The last step for global is to select the only recods where cases are greater then zero.

```{r}
global <- global %>% filter(cases > 0)
summary(global)
```

##### Let's check see what data displayed

```{r}
global %>% filter(cases > 28000000)
```

##### Now we repeat the same procedure as above for the US_cases and US_deaths

```{r}
US_cases <- US_cases %>%
    pivot_longer(cols = -(UID:Combined_Key),
                 names_to = "date",
                 values_to = "cases") %>%
    select(Admin2:cases) %>%
    mutate(date = mdy(date)) %>%
    select(-c(Lat, Long_))
US_cases      

US_deaths <- US_deaths %>%
    pivot_longer(cols = -(UID:Population),
                 names_to = "date",
                 values_to = "deaths") %>%
    select(Admin2:deaths) %>%
    mutate(date = mdy(date)) %>%
    select(-c(Lat, Long_))
US_deaths

US <- US_cases %>% full_join(US_deaths)
```

##### The difference between US and global is that global do not have poplulaton column. Thus, I will download extra table which contains population column and add it to global table.

```{r}
global <- global %>%
    unite("Combined_Key",
          c(Province_State, Country_Region),
          sep = ", ",
          na.rm = TRUE,
          remove = FALSE)
global

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
    select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
    left_join(uid, by = c("Province_State", "Country_Region")) %>%
    select(-c(UID, FIPS)) %>%
    select(Province_State, Country_Region, date,
           cases, deaths, Population,
           Combined_Key)
global
```

# Visualizing Data
##### Let's visualize the data that shows number of cases and deaths per date in each country regiion

```{r}
US_by_state <- US %>%
    group_by(Province_State, Country_Region, date) %>%
    summarise(cases = sum(cases), deaths = sum(deaths),
              Population = sum(Population)) %>%
    mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
    select(Province_State, Country_Region, date,
           cases, deaths, deaths_per_mill, Population) %>%
    ungroup()
US_by_state

US_totals <- US_by_state %>%
    group_by(Country_Region, date) %>%
    summarise(cases = sum(cases), deaths = sum(deaths),
              Population = sum(Population)) %>%
    mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
    select(Country_Region, date,
           cases, deaths, deaths_per_mill, Population) %>%
    ungroup()
US_totals
tail(US_totals)

US_totals %>%
    ggplot(aes(x = date, y = cases)) +
    geom_line(aes(color = "cases")) +
    geom_point(aes(color = "cases")) +
    geom_line(aes(y = deaths, color = "deaths")) +
    geom_point(aes(y = deaths, color = "deaths")) +
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
    labs(title = "COVID19 in US", y = NULL)
```

##### Next let's see the result just for California state

```{r}
state <- "California"
US_by_state %>%
    filter(Province_State == state) %>%
    filter(cases > 0) %>%
    ggplot(aes(x = date, y = cases)) +
    geom_line(aes(color = "cases")) +
    geom_point(aes(color = "cases")) +
    geom_line(aes(y = deaths, color = "deaths")) +
    geom_point(aes(y = deaths, color = "deaths")) +
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
    labs(title = str_c("COVID19 in ", state), y = NULL)
max(US_totals$deaths)
```

# Analyzing Data
##### Let's see the relationship between US covid-19 cases and deaths
```{r}
US_by_state <- US_by_state %>%
    mutate(new_cases = cases - lag(cases),
           new_deaths = deaths -lag(deaths))
US_totals <- US_totals %>%
    mutate(new_cases = cases - lag(cases),
           new_deaths = deaths -lag(deaths))
tail(US_totals)
tail(US_totals %>% select(new_cases, new_deaths, everything()))


US_totals %>%
    ggplot(aes(x = date, y = new_cases)) +
    geom_line(aes(color = "new_cases")) +
    geom_point(aes(color = "new_cases")) +
    geom_line(aes(y = new_deaths, color = "new_deaths")) +
    geom_point(aes(y = new_deaths, color = "new_deaths")) +
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
    labs(title = "COVID19 in US", y = NULL)
```

##### The above graph show that number of cases is proportional to number of deaths. 
##### Let's see how number of cases and deaths interact in California

```{r}
state <- "California"
US_by_state %>%
    filter(Province_State == state) %>%
    filter(cases > 0) %>%
    ggplot(aes(x = date, y = new_cases)) +
    geom_line(aes(color = "new_cases")) +
    geom_point(aes(color = "new_cases")) +
    geom_line(aes(y = new_deaths, color = "new_deaths")) +
    geom_point(aes(y = new_deaths, color = "new_deaths")) +
    scale_y_log10() + 
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90)) +
    labs(title = str_c("COVID19 in ", state), y = NULL)

US_state_totals <- US_by_state %>%
    group_by(Province_State) %>%
    summarise(deaths = max(deaths), cases = max(cases),
              population = max(Population),
              cases_per_thou = 1000 * cases / population,
              deaths_per_thou = 1000 * deaths / population) %>%
    filter(cases > 0, population > 0)

US_state_totals %>%
    slice_min(deaths_per_thou, n = 10) %>%
select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>%
    slice_max(deaths_per_thou, n = 10) %>%
select(deaths_per_thou, cases_per_thou, everything())
```

##### The result for California is very similar to the result of the USA.
# Modeling Data
##### To see a better picture I would like to see correlation between deaths an cases.

```{r}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)

US_state_totals %>% slice_min(cases_per_thou)
US_state_totals %>% slice_max(cases_per_thou)
US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred

US_tot_w_pred %>% ggplot() +
    geom_point(aes(x = cases_per_thou, y = deaths_per_thou), 
               color = "blue") +
    geom_point(aes(x = cases_per_thou, y = pred),
               color = "red")
```

##### The graph above shows that prediction of number of deaths based on number of cases.

# Conclusion and Bias
The analysis above shows that the number of cases plays a primary role in the number of deaths, although some points are far away from prediction. The bias of this analysis could be that it is very questionable if covid-19 caused the deaths or some other factors. Many people were tested positive, but not so many died. The deaths collected in the data source may be bais because it is possible that not Covid-19 played the primary role for the death but some prior condition of the body.